---
title: "DA2 Homework Classification 1"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Get *[Quote_ID],[Quote], [Competitor_Quote] and [Result] FROM [dbo].[Quote]* in the Accouting database. Then, create a column named QuoteDiff. Review a density plot of QuoteDiff based on Result.

```{r, message=F, warning=F, fig.width=5, fig.height=3, fig.align="center"}

library(DBI)
library(tidyverse)
library(odbc)
library(stringr)
library(DBI)
library(MASS)
library(ROCR)
library(lubridate)
library(DMwR)
library(caret)


con2 <- dbConnect(odbc::odbc(), Driver = "{SQL Server}", Server = "analyticslab.database.windows.net", Port = "1433", Database = "Accounting", UID = "Student", 
    PWD = "Acct7397")

QuoteData <- dbGetQuery(con2,"
SELECT [dbo].[Quote].[Quote_ID]
                         ,CAST([dbo].[Quote].[Quote] AS FLOAT) AS Quote
                         ,CAST([dbo].[Quote].[Competitor_Quote] AS FLOAT) AS Competitor_Quote
                         ,[dbo].[Quote].[Result] 
                         FROM [dbo].[Quote] 
                         ")


QuoteData$Quote <-as.numeric (QuoteData$Quote)
QuoteData$Competitor_Quote <-as.numeric (QuoteData$Competitor_Quote)
QuoteData$QuoteDiff <- as.numeric(QuoteData$Competitor_Quote - QuoteData$Quote)

QuoteData$Result <- as.factor(QuoteData$Result)
QuoteData$Result[QuoteData$Result=="l"] <- "L" # fix a data error
QuoteData$Result <- factor(QuoteData$Result)

pl1 <- ggplot(QuoteData, aes(QuoteDiff, fill = Result)) 
pl1 <- pl1 + geom_density(alpha = 0.2, adjust = 5 )
pl1 

set.seed(9164)


testSplit <- .4
totalSampleSize <- nrow(QuoteData)
testSampleSize <- round(totalSampleSize*testSplit)
trainSampleSize <- totalSampleSize - testSampleSize
tindexes <- sample(1:nrow(QuoteData), testSampleSize)
indexes <- sample(1:nrow(QuoteData[-tindexes,]), trainSampleSize)
xTrain <- QuoteData[indexes, ]
xTest <- QuoteData[tindexes,]


lda.fit <- lda(Result ~ QuoteDiff, xTrain) 
lda.pred <- predict(lda.fit, xTest)




A <- A <- mean(lda.fit$means)
B <- log(lda.fit$prior[2]) - log(lda.fit$prior[1])
s2.k <- t(tapply(xTest$QuoteDiff, xTest$Result, var)) %*% lda.fit$prior
C <- s2.k/(lda.fit$means[1] - lda.fit$means[2])
dr <- A + B * C

pl1 <- pl1 + geom_vline(xintercept = dr, color = "red" )
pl1

```
Build a model to predict Result (W or L) based on the QuoteDiff only. Use LDA. Create a confusition matrix, and show. Should be similar to the following:


```{r, message=F, warning=F, fig.width=6, fig.height=4, fig.align="center"}

confusionMatrix(factor(lda.pred$class), factor(xTest$Result), positive = "W")

```

Now plot out the ROC curve.

```{r, message=F, warning=F, fig.width=4, fig.height=4, fig.align="center"}


tab <- table(lda.pred$class, xTest$Result, 
             dnn = c('Predicted', 'Actual')) 

addmargins(tab)

sens <- tab[2,2]/(tab[1,2] + tab[2,2]); 
spec <- tab[1,1]/(tab[1,1] + tab[2,1]); 

str_c("Sensitivity = ", sens)
str_c("Specificity = ", spec)

# ROC Curve
yTest <- xTest$Result
pred <- prediction(lda.pred$posterior[,2], xTest$Result) 
perf <- performance(pred, measure = 'tpr', x.measure = 'fpr')
plot(perf)
auc<- performance(pred,"auc")
str_c("AUC = ", auc@y.values)


```

Now, show model predictions on a plot as shown below:

```{r, message=F, warning=F, fig.width=6, fig.height=4, fig.align="center"}


ggplot(data = QuoteData, aes(x= QuoteDiff, y = Quote_ID, color = Result)) + 
  geom_point() +    
  theme(panel.background = element_rect(fill = "white")) 


```


